The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015

### 2017-09-11 paper/paper.tex
Fixed most graph/table positons - ran everything, so far good

% Try plotting this for each country and wave
% Like, every country as a row and 6 columns for each wave.
\begin{figure}
\begin{center}

@@

\caption{Standard deviation of test scores across all waves.}
\end{center}
\end{figure}

@@

% New table only has sample size, average score and s.e.
% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Wed Jul  5 17:51:02 2017
\scalebox{0.75}{
\begin{tabular}{lllrrlrr}
   \hline \
                          & & & Low SES & & & High SES \\
                          \cmidrule(l){3-5}
                          \cmidrule(l){6-8} Year & Countries & N & Avg score & S.E & N & Avg score & S.E \\ \hline2000 & Australia & 294 & 0.22 & 0.07 & 270 & 1.22 & 0.09 \\ 
  2000 & Austria & 294 & -0.04 & 0.11 & 235 & 1.15 & 0.15 \\ 
  2000 & Canada & 1970 & 0.30 & 0.05 & 1357 & 0.90 & 0.07 \\ 
  2000 & Denmark & 230 & -0.01 & 0.13 & 228 & 1.17 & 0.17 \\ 
  2000 & Finland & 261 & 0.42 & 0.12 & 267 & 0.89 & 0.15 \\ 
  2000 & France & 253 & 0.04 & 0.04 & 258 & 1.21 & 0.05 \\ 
  2000 & Germany & 250 & -0.64 & 0.04 & 303 & 1.81 & 0.05 \\ 
  2000 & Hungary & 291 & -0.46 & 0.09 & 254 & 1.66 & 0.12 \\ 
  2000 & Italy & 278 & -0.31 & 0.05 & 278 & 0.81 & 0.06 \\ 
  2000 & Netherlands & 129 & 0.46 & 0.08 & 158 & 1.25 & 0.11 \\ 
  2000 & Poland & 209 & -0.31 & 0.04 & 166 & 1.25 & 0.06 \\ 
  2000 & Spain & 318 & -0.27 & 0.05 & 340 & 1.09 & 0.07 \\ 
  2000 & Sweden & 241 & 0.12 & 0.10 & 246 & 1.06 & 0.13 \\ 
  2000 & United Kingdom & 531 & 0.16 & 0.04 & 415 & 1.33 & 0.06 \\ 
  2000 & United States & 222 & -0.43 & 0.02 & 164 & 1.74 & 0.03 \\ 
  2015 & Australia & 1694 & -0.15 & 0.04 & 1235 & 1.28 & 0.05 \\ 
  2015 & Austria & 676 & -0.30 & 0.07 & 705 & 1.53 & 0.10 \\ 
  2015 & Canada & 2215 & 0.20 & 0.03 & 1863 & 0.99 & 0.05 \\ 
  2015 & Denmark & 1013 & 0.09 & 0.08 & 595 & 1.08 & 0.10 \\ 
  2015 & Finland & 575 & 0.07 & 0.08 & 584 & 1.16 & 0.11 \\ 
  2015 & France & 570 & -0.37 & 0.02 & 615 & 1.66 & 0.03 \\ 
  2015 & Germany & 545 & -0.09 & 0.02 & 582 & 1.45 & 0.03 \\ 
  2015 & Hungary & 466 & -0.72 & 0.07 & 589 & 1.79 & 0.09 \\ 
  2015 & Italy & 947 & -0.20 & 0.03 & 1043 & 1.23 & 0.04 \\ 
  2015 & Netherlands & 521 & 0.13 & 0.04 & 525 & 1.24 & 0.06 \\ 
  2015 & Poland & 446 & 0.10 & 0.03 & 448 & 1.24 & 0.05 \\ 
  2015 & Spain & 608 & -0.25 & 0.03 & 703 & 1.24 & 0.04 \\ 
  2015 & Sweden & 527 & -0.20 & 0.06 & 542 & 1.32 & 0.09 \\ 
  2015 & United Kingdom & 1387 & -0.03 & 0.03 & 1388 & 1.15 & 0.04 \\ 
  2015 & United States & 585 & -0.39 & 0.01 & 544 & 1.22 & 0.01 \\ 
   \hline
\end{tabular}
}

@@

We see that in some countries like Finland and Sweden the average low SES scores is actually above the average score of 0, suggesting their very equal countries. However, most countries, like Spain, Poland, Hungary and Germany, their average low SES scores are much lower than the country average, which is 0. We also see some countries with major changes from year 2000 to 2015, with, for example, Australia decreasing the low SES gap from 0.22 standard deviations to -0.15, well below the country average. In the next section we take a look at this in a more detailed fashion, considering all years in between.

\subsection{Evolution of the achievement gap}
% TODO: Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).

\begin{figure}
\begin{center}

@@

\caption{Evolution of the achievement gap for selected countries}
\end{center}
\end{figure}

We start by look at the achievement for some countries in figure 2. We plot only mathematics for each country and also a quadratic trend spline for \emph{both mathematics and literacy} pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden have very steep slopes. France increasde the gap by roughly 0.9 standard deviations, Austria by 0.6 and Sweden by 0.6. This pattern happens similarly for literacy. For example, France has an increase of 0.6. For such a short period of time, the magnitud of these increases are reasonably big.

@@

Other countries have reasonable increases such as Finland and Hungary, with increases of nearly 0.6 and 0.4, respectively. Aside from these countries, there are other cases that have no change at all, specifically, Canada, Netherlands and Spain. Canada excels here not only because the gap has been stable over time, but because it has the smallest gap of all countries presented here. It's nearly 0.5 standard deviations in 2000 and it increased only by 0.2 in 2015. We can't really say Italy has had a significant since it has a sort of wiggly pattern that looks like it cancels itself out.

On the other hand, we do have some countries which show a decrease in the SES achievement gap. United Kingdom decreased by about 0, Poland by -0.4, and Denmark by -0.2. However, the most notable cases are the United States and Germany. These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations. But in the 15-year time trend both countries reduced the gaps by -0.6 and -0.9 respectively. Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}. This corroborates the findings of \citet{reardon_portilla}, which found a decreasing gap for kindergarteners. This decline is evident also for 15 year olds, suggesting it might be more of an institutional change rather than a specific grade-level policy.

Analyzing this graph we might naively conclude that these trends are not very steep and they should not be very important to consider. However, remember that the Y axis is measured in standard deviations. Small changes are actually very huge in practical terms. Take the case of Sweden. The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015. The actual difference is of about 0.6. Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.

We find that the initial gap for the U.S in 2000 is 2.17 varying between 2.13 and 2.21. This gap is much higher than found in previous studies, but again, we don't know much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely. All country-level studies pool many studies together and control for age, removing the age-specific gap effect.

@@

\begin{figure}
\begin{center}

@@

\caption{Percentage change in the achievement gap from 2000 to 2015}
\end{center}
\end{figure}

Generally speaking, we see that most countries increased their achievement gap over time. France had an average increase of about 80\% since 2000 varying down to 40\%, whereas Germany had a similar figure but decreasing. Many of the countries we saw before that didn't have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap. In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%. The benefit of presenting these estimates this way is that we can actually asses the uncertainty of each calculation, and we do find that some of these estimates have wide variability. Despite this, most countries show a clear sign of either decreasing or increasing.

@@

% Add confidence intervals
\begin{figure}
\begin{center}

@@

\caption{Evolution of the achievement gap for separate SES groups}
\end{center}
\end{figure}

Figure 4 shows the evolution for both Low and High SES groups separetly. The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other. We find plenty of variation in these patterns. For example, in the United States the top seems to be equalizing much stronger than than the bottom is catching up. The U.K seems to be following the same path as the United States as well. On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster. The Netherlands shows a similar pattern as the U.S, U.K patterns but upon closer inspection the explanation is very different. We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.

Finally, we pay attention to the flat summary lines. These lines should be interpreted with caution because it doesn't mean that the gap is not increasing. Denmark, Belgium, Netherlands, Norway and Spain show a flat line because the gap is growing very little, if at all. On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate. These results highlight the importance of not only summarizing average achievement gaps. The source of these gaps vary greatly between countries and we see how each of these patterns contibutes to the overall inequality of a country.

Ths source of this gap can be deconstructed further into two broader gaps. We want to know if these gaps have increased because the 90th SES group has distanced itself from the 50th SES group or because the middle has been also distancing itself from 10th SES group. For example, we're interested in searching which gap is bigger between countries, the 90th/50th or the 50th/10th? Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier? Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap in figure 5. Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous graphs \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.

\begin{figure}
\begin{center}

@@

\caption{Evolution of 90/50/10 SES gaps for selected countries}
\end{center}
\end{figure}

@@

\caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
\end{center}
\end{figure}

We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as the United States and Germany. We see a negative relationship where the faster the low SES catch up the slower the High SES decrease. This is very interesting because it suggests that there's a trend for countries to have the bottom SES groups are climbing upwards, a trait of meritocracy. Equally important, we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more ocuntries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

@@

\caption{Evolution of the SES gaps for top/bottom groups}
\end{center}
\end{figure}

@@

Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's a tradeoff between the average country performance and the widening of the achievement gap and if it stands over time. One particular topic that has been researched in the past is whether improving the overall performance of a country compromises the equality of the educational system. For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance. This, however, can come at a price. If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance. We plot this relationship in figure 8.

\subsection{Performance-equality tradeoff}

\begin{figure}
\begin{center}

@@

\caption{Relationship between 90/10 achievement gap and average performance in mathematics}
\end{center}
\end{figure}

For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}. The correlations are between -0.73 and -0.88. This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}. However, what they didn't really go in depth was whether this corelation is strongest for some countries over time. With the data available we can look at the evolution separately for many countries. We plot it up next in figure 9.

% Add confidence intervals
\begin{figure}
\begin{center}

@@

\caption{Time trends of achievement gap and average achievement between 2000 and 2015}
\end{center}
\end{figure}

@@

\section{The role of curricular tracking in explaining inequalities}

@@

In the final model, we regress the inequality of a country on a dummy variable where 1 equals only 1 track vs more than 1 track, whether the age of first selection is 15 or more vs below 15, a dummy stating whether the percentage of tracked curriculum is above 1, and the standardized vocational index. We are reluctant to include more variable in the model, first to keep it more parsimonious, and secondly to prevent overfitting due to our sample size. We present the results in the table below.

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Sep  7 15:04:53 2017
\begin{table}[ht]
\centering
\begin{tabular}{lllll}
  \hline
  & Model 1 & Model 2 & Model 3 & Model 4 \\ 
  \hline
Only 1 track & -0.53 (-0.64/-0.41) & -0.31 (-0.44/-0.18) & -0.31 (-0.45/-0.17) & -0.21 (-0.34/-0.07) \\ 
  Age selection $>$= 15 &  & -0.38 (-0.51/-0.25) & -0.38 (-0.51/-0.26) & -0.57 (-0.71/-0.43) \\ 
  \% of curric tracked &  &  & 0 (-0.23/0.23) & 0.6 (0.3/0.89) \\ 
  Vocational Index &  &  &  & -0.27 (-0.36/-0.19) \\ 
  Intercept & 1.77 (1.69/1.84) & 1.94 (1.85/2.03) & 1.94 (1.69/2.19) & 1.54 (1.27/1.81) \\ 
   \hline R square: & 23\% & 32\% & 32\% & 40\% \\ Sample size: & 195 & 195 & 195 & 183 \\ \hline \hline
\end{tabular}
\end{table}

Given that the classical p values are inexistent in bayesian modelling, I present uncertainty intervals next to each coefficient. The first model only includes the 1 track dummy. This variable shows that countries that have one track, have on average, half a standard deviation less inequality than other countries. This covariate alone explains about 23\% of the variance of the evolution of the 90/10 achievement gap. The second model now includes the dummy of whether the country has an age selection of 15 or more and we can see that countries with late selection into tracking have about -.38 less standard deviations in inequality. Also note that the number of tracks dummy was also reduced, suggesting these two variables are explaining similar things. This is expected given that late tracking is usually associated with fewer tracks, such as in the scandinavian countries. The third model now includes whether the country has any degree of tracking, meaning whether the country has above 0\% of the curriculum tracked. The coding of the variable comes from visual inspection of the distribution, where some countries, such as the United States and Canada, have untracked curriculums. We see now relationship at all, but after including the vocational index, the model is completely reversed with countries with tracked curriculum having about 0.6 more standard deviations in inequality that countries with no tracking.

We also note that vocational enrollment is associated with less inequality. Not surprisingly, the age of selection coefficient, also increased due to the correlation between age of selection and vocational enrollment. Countries with high vocational enrollment are also those with early age selection, such as Germany and Austria. These two variables should be inspected further, perhaps with an interaction, given than both coefficients are sizable and very correlated. This final model explains over one third of the variance in the evolution of the 90/10 SES gap, with an R square of 40\%. The uncertainty of each of the covariates varies a lot with the percentage of curriculum tracked dummy ranging all the way from 0.9 to 0.3. Nonetheless, we see that the other coefficients are very strong and have reasonable intervals. It is very important to consider the uncertainty of our estimates, to give confidence in our model fit. So far, this model seems to fit the data quite reasonably. \footnote{These variables are very correlated, but no collinearity is present with a maximum VIF of 2.2.}

All of these countries have a different tracking structure, with some setups being more egalitarian than others. Using the previous model we make a simulation and predict the level of the 90/10 achievement gap for all possible tracking setups for all countries. Below we plot all these combinations. The x axis has the number of tracks that a country has, the y axis has the predicted 90/10 achievement gap and below each country name we have whether the country has an age selection of 15 or more or below. In addition, to confirm how accurate is the model, we also plot the actual level of inequality in the actual tracking setup of the country. Let's take Germany for example.

>>>>

% Try plotting this for each country and wave
% Like, every country as a row and 6 columns for each wave.
% \begin{figure}
% \begin{center}
\begin{figure}

@@

\caption[Standard deviation of test scores across all waves]{Standard deviation of test scores across all waves}\label{fig:variance_pisa}
\end{figure}

% \caption{Standard deviation of test scores across all waves}
% \end{center}
% \end{figure}

@@

% New table only has sample size, average score and s.e.
% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Sat Sep  9 20:13:40 2017
\begin{table}[ht]
\centering
\begin{tabular}{lllrrlrr}
   \hline \
                          & & & Low SES & & & High SES \\
                          \cmidrule(l){3-5}
                          \cmidrule(l){6-8} Year & Countries & N & Avg score & S.E & N & Avg score & S.E \\ \hline2000 & Australia & 294 & 0.22 & 0.07 & 270 & 1.22 & 0.09 \\ 
  2000 & Austria & 294 & -0.04 & 0.11 & 235 & 1.15 & 0.15 \\ 
  2000 & Canada & 1970 & 0.30 & 0.05 & 1357 & 0.90 & 0.07 \\ 
  2000 & Denmark & 230 & -0.01 & 0.13 & 228 & 1.17 & 0.17 \\ 
  2000 & Finland & 261 & 0.42 & 0.12 & 267 & 0.89 & 0.15 \\ 
  2000 & France & 253 & 0.04 & 0.04 & 258 & 1.21 & 0.05 \\ 
  2000 & Germany & 250 & -0.64 & 0.04 & 303 & 1.81 & 0.05 \\ 
  2000 & Hungary & 291 & -0.46 & 0.09 & 254 & 1.66 & 0.12 \\ 
  2000 & Italy & 278 & -0.31 & 0.05 & 278 & 0.81 & 0.06 \\ 
  2000 & Netherlands & 129 & 0.46 & 0.08 & 158 & 1.25 & 0.11 \\ 
  2000 & Poland & 209 & -0.31 & 0.04 & 166 & 1.25 & 0.06 \\ 
  2000 & Spain & 318 & -0.27 & 0.05 & 340 & 1.09 & 0.07 \\ 
  2000 & Sweden & 241 & 0.12 & 0.10 & 246 & 1.06 & 0.13 \\ 
  2000 & United Kingdom & 531 & 0.16 & 0.04 & 415 & 1.33 & 0.06 \\ 
  2000 & United States & 222 & -0.43 & 0.02 & 164 & 1.74 & 0.03 \\ 
  2015 & Australia & 1694 & -0.15 & 0.04 & 1235 & 1.28 & 0.05 \\ 
  2015 & Austria & 676 & -0.30 & 0.07 & 705 & 1.53 & 0.10 \\ 
  2015 & Canada & 2215 & 0.20 & 0.03 & 1863 & 0.99 & 0.05 \\ 
  2015 & Denmark & 1013 & 0.09 & 0.08 & 595 & 1.08 & 0.10 \\ 
  2015 & Finland & 575 & 0.07 & 0.08 & 584 & 1.16 & 0.11 \\ 
  2015 & France & 570 & -0.37 & 0.02 & 615 & 1.66 & 0.03 \\ 
  2015 & Germany & 545 & -0.09 & 0.02 & 582 & 1.45 & 0.03 \\ 
  2015 & Hungary & 466 & -0.72 & 0.07 & 589 & 1.79 & 0.09 \\ 
  2015 & Italy & 947 & -0.20 & 0.03 & 1043 & 1.23 & 0.04 \\ 
  2015 & Netherlands & 521 & 0.13 & 0.04 & 525 & 1.24 & 0.06 \\ 
  2015 & Poland & 446 & 0.10 & 0.03 & 448 & 1.24 & 0.05 \\ 
  2015 & Spain & 608 & -0.25 & 0.03 & 703 & 1.24 & 0.04 \\ 
  2015 & Sweden & 527 & -0.20 & 0.06 & 542 & 1.32 & 0.09 \\ 
  2015 & United Kingdom & 1387 & -0.03 & 0.03 & 1388 & 1.15 & 0.04 \\ 
  2015 & United States & 585 & -0.39 & 0.01 & 544 & 1.22 & 0.01 \\ 
   \hline
\end{tabular}
\caption{SES sample size and ISCED composition} 
\end{table}

@@

We see that in some countries like Finland and Sweden the average low SES scores is actually above the average score of 0, suggesting they're very equal countries. However, most countries, like Spain, Poland, Hungary and Germany, their average low SES scores are much lower than the country average, which is 0. We also see some countries with major changes from year 2000 to 2015, with, for example, Australia decreasing the low SES gap from 0.22 standard deviations to -0.15, well below the country average. In the next section we take a look at this in a more detailed fashion, considering all years in between.

% TODO: Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).

% \begin{figure}
% \begin{center}
\begin{figure}

@@

\caption[Evolution of the achievement gap for selected countries]{Evolution of the achievement gap for selected countries}\label{fig:graphing_9010gaps}
\end{figure}

% \caption{Evolution of the achievement gap for selected countries}
% \end{center}
% \end{figure}

We start by look at the achievement for some countries in figure 2. We plot only mathematics for each country and also a quadratic trend spline for \emph{both mathematics and literacy} pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden have very steep slopes. France increased the gap by roughly 0.9 standard deviations, Austria by 0.6 and Sweden by 0.6. This pattern happens similarly for literacy. For example, France has an increase of 0.6. For such a short period of time, the magnitud of these increases are reasonably big.

@@

Other countries have reasonable increases such as Finland and Hungary, with increases of nearly 0.6 and 0.4, respectively. Aside from these countries, there are other cases that have no change at all, specifically, Canada, Netherlands and Spain. Canada excels here not only because the gap has been stable over time, but because it has the smallest gap of all countries presented here. It's nearly 0.5 standard deviations in 2000 and it increased only by 0.2 in 2015. We can't really say Italy has had a significant increase since it has a sort of wiggly pattern that looks like it cancels itself out.

On the other hand, we do have some countries which show a decrease in the SES achievement gap. Poland decreased by about -0.4 and Denmark by -0.2. However, the most notable cases are the United States and Germany. These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations. But in the 15-year time trend both countries reduced the gaps by -0.6 and -0.9 respectively. Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}. This corroborates the findings of \citet{reardon_portilla}, which found a decreasing gap for kindergarteners. This decline is evident also for 15 year olds, suggesting it might be more of an institutional change rather than a specific grade-level policy. Do note that the trendline is for both mathematics and literacy. If it were only for mathematics, then the trendline would probably be even more extreme. Despite this, pooling both subjects gives even more strength to the analysis given that the results stand for both subjects together \footnote{Same graph is availabe for reading upon request, where we find pretty much the same results if not even more steep/downward slopes.}. We also plot the 80/20 and 70/30 SES gaps and find that very similar patterns as in the 90/10 SES gap (see figure 10 in the appendix).

Analyzing this graph we might naively conclude that these trends are not very steep and they should not be very important to consider. However, remember that the Y axis is measured in standard deviations. Small changes are actually very huge in practical terms. Take the case of Sweden. The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015. Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.

We find that the initial gap for the U.S in 2000 is 2.17 varying between 2.13 and 2.21. This gap is much higher than found in previous studies, but again, we don't know how much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely. All country-level studies pool many studies together and control for age, removing the age-specific gap effect.

@@

% \begin{figure}
% \begin{center}
\begin{figure}

@@

\caption[Evolution of the achievement gap for separate SES groups]{Evolution of the achievement gap for separate SES groups}\label{fig:graphing_perc_increase}
\end{figure}

% \caption{Percentage change in the achievement gap from 2000 to 2015}
% \end{center}
% \end{figure}

Generally speaking, we see that most countries increased their achievement gap over time. France had an average increase of about 80\% since 2000 varying down to 40\%, whereas Germany had a similar figure but decreasing. Many of the countries we saw before that didn't have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap. In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%. The benefit of presenting these estimates this way is that we can actually asses the uncertainty of each calculation, and we do find that some of these estimates have wide variability. Despite this, most countries show a clear sign of either decreasing or increasing. We plot the same graph for the 80/20 and 70/30 SES gaps in the appendix and the results hold for these achievement gaps as well.

@@

% Add confidence intervals
% \begin{figure}
% \begin{center}
\begin{figure}

@@

\caption[Evolution of the achievement gap for separate SES groups]{Evolution of the achievement gap for separate SES groups}\label{fig:rate_increase/decrease}
\end{figure}

% \caption{Evolution of the achievement gap for separate SES groups}
% \end{center}
% \end{figure}

Figure 4 shows the evolution for both low and high SES groups separately. The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other. We find plenty of variation in these patterns. For example, in the United States the top seems to be equalizing much stronger than than the bottom is catching up. The U.K seems to be following the same path as the United States as well. On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster. The Netherlands shows a similar pattern as the U.S, U.K patterns but upon closer inspection the explanation is very different. We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.

We now pay attention to the flat summary lines. These lines should be interpreted with caution because it doesn't mean that the gap is not increasing. Denmark, Belgium, Netherlands, Norway and Spain show a flat line because the gap is growing very little, if at all. On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate. These results highlight the importance of not only summarizing average achievement gaps. The source of these gaps vary greatly between countries and we see how each of these patterns contibutes to the overall inequality of a country.

After analyzing the trend for each country and where this trend is coming from we start to find some interesting results. We see that the United States is closing the achievement gap at a very rapid rate, but at the same time we notice that it's because the top SES group is going down faster than the bottom is catching up. In contrast, we see countries like Germany and Poland where the gap is closing as well but because bottom group is catching up faster than the high SES group is coming down. In fact, the Polish case is the most egalitarian of all the countries presented here; the bottom group is catching up at a rapid pace and the top group is maintaining their high levels of performance, fostering a trully equalizing effect.

Quite interestingly, in countries where the gap is widening, both groups are distancing themselves at a very similar pace. Finland, France and Austria, show this exact pattern. Give or take, we have a solid pattern that when the gap widens, it's because there's top-to-bottom inequality. But when the gap is being reduced, the source can vary between top and bottom changes.

To help assess these results even further, the source of this gap can be deconstructed further into two broader gaps. We want to know if these gaps have increased because the 90th SES group has distanced itself from the 50th SES group or because the middle has been also distancing itself from 10th SES group. For example, we're interested in searching which gap is bigger between countries, the 90th/50th or the 50th/10th? Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier? Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap in figure 5. Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous graphs \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.

% \begin{figure}
% \begin{center}
\begin{figure}

@@

\caption[Evolution of 90/50/10 SES gaps for selected countries]{Evolution of 90/50/10 SES gaps for selected countries}\label{fig:graphing_allgaps}
\end{figure}

% \caption{Evolution of 90/50/10 SES gaps for selected countries}
% \end{center}
% \end{figure}

@@

\caption[Rate at which top/bottom SES groups are catching up for the 90/10 SES gap]{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}\label{fig:rate_change_graph}
\end{figure}

% \caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
% \end{center}
% \end{figure}

We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (top-left panel) but there's also a fair share where low SES is catching up faster, such as the United States and Germany. Regardless of this, we see an overall negative relationship where countries in which the low SES group is catching up, the high SES group goes down. This is very interesting because it suggests that there's a trend for countries to have the bottom SES groups are climbing upwards, a trait of meritocracy. However, this means that the top group comes down, rather than stagnate or increase. Equally important, we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more countries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

To sum up, we find no evidence in support of our first hypothesis. We uncover that the tracking setup of a country doesn't imminently mean a certain degree of equality/inequality. Countries where the top SES group is increasing and the bottom is decreasing (top-left panel) have dramatically different tracking setups. Austria is an early tracker with several tracks, while Sweden is a late tracker with only 1 track. On the opossite panel (bottom-right), where we find countries that the bottom group is catching up while the top is coming down, we find similar results. Denmark and Germany are opposite poles in terms of tracking, while the U.S is distinctive in that it has little formal curricular stratification. We see no definitive evidence of top-to-bottom equality in all of these countries (except for Poland). We find that regardless of the different tracking setups, we see that there's a tradeoff: either the bottom comes up and the top comes down or the top goes even further up and the bottom further down.

@@

\caption[Evolution of the SES gaps for top/bottom groups]{Evolution of the SES gaps for top/bottom groups}\label{fig:graphing_ses_growth}
\end{figure}

% \caption{Evolution of the SES gaps for top/bottom groups}
% \end{center}
% \end{figure}

@@

\subsection{Equality-efficiency tradeoff}

Moving on, we now discuss if there's a tradeoff between an educational system being efficient, i.e raising the overall performance of the country and whether that challenge compromises equality at the same time. For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance. This, however, can come at a price. If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance. Given that we have the data for the average performance for each country and the achievement gap, we can investigate this relationship in detail. We first being by plotting all countries pooled for each year in figure 8.

% \begin{figure}
% \begin{center}
\begin{figure}

@@

\caption[Relationship between 90/10 achievement gap and average performance in mathematics]{Relationship between 90/10 achievement gap and average performance in mathematics}\label{fig:avgperformance_inequality}
\end{figure}

% \caption{Relationship between 90/10 achievement gap and average performance in mathematics}
% \end{center}
% \end{figure}

For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}. The correlations are between -0.73 and -0.88. This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}. However, what they didn't really go in depth was whether this corelation is strongest for some countries over time. However, we should not get trapped into an ecological fallacy and assume this relationship is identical within each country. With the data available we can look at the evolution separately for many countries. We plot it up next in figure 9.

% Add confidence intervals
% \begin{figure}
% \begin{center}
\begin{figure}

@@

\caption[Time trends of achievement gap and average achievement between 2000 and 2015]{Time trends of achievement gap and average achievement between 2000 and 2015}\label{fig:country_achievement_and_inequality}
\end{figure}

% \caption{Time trends of achievement gap and average achievement between 2000 and 2015}
% \end{center}
% \end{figure}

@@

These results actually confirm our second hypothesis, that there's a negative relationship between the two variables. We see countries like Poland, Canada and Germany where the pattern is just as the aggregate trend, namely that as the average performance increases, inequality seems to be reduced in parallel. Conversely, France, Austria and Australia, to name a few, portray the opposite: as average performance decreases inequality increases. But we also unconver important variation within countries in the relationship between average achievement and dispersion of achievement (have a look at the Netherlands, Italy and the U.S).

\subsection{The role of curricular tracking in explaining inequalities}

@@

In the final model, we regress the inequality of a country on a dummy variable where 1 equals only 1 track vs more than 1 track, whether the age of first selection is 15 or more vs below 15, a dummy stating whether the percentage of tracked curriculum is above 0 (that is, any tracking), and the standardized vocational index \footnote{We create dummies to avoid multicollinearity. The final model has a maximum VIF of 2.2.}. We are reluctant to include more variable in the model, first to keep it more parsimonious, and secondly to prevent overfitting due to our sample size. We present the results in the table below.

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Sun Sep 10 14:55:49 2017
\begin{table}[ht]
\centering
\begin{tabular}{lllll}
  \hline
  & Model 1 & Model 2 & Model 3 & Model 4 \\ 
  \hline
Only 1 track & -0.53 (-0.64/-0.41) & -0.31 (-0.44/-0.18) & -0.31 (-0.45/-0.17) & -0.21 (-0.34/-0.07) \\ 
  Age selection $>$= 15 &  & -0.38 (-0.51/-0.25) & -0.38 (-0.51/-0.26) & -0.57 (-0.71/-0.43) \\ 
  \% of curric tracked &  &  & 0 (-0.23/0.23) & 0.6 (0.3/0.89) \\ 
  Vocational Index &  &  &  & -0.27 (-0.36/-0.19) \\ 
  Intercept & 1.77 (1.69/1.84) & 1.94 (1.85/2.03) & 1.94 (1.69/2.19) & 1.54 (1.27/1.81) \\ 
   \hline R square: & 23\% & 32\% & 32\% & 40\% \\ Sample size: & 195 & 195 & 195 & 183 \\ \hline \hline
\end{tabular}
\caption{Explaining 90/10 achievement gap - Multilevel model} 
\end{table}

Given that the classical p values are inexistent in bayesian modelling, we present uncertainty intervals next to each coefficient. The first model only includes the 1 track dummy. This variable shows that countries that have one track, have on average, half a standard deviation less inequality than other countries. This covariate alone explains about 23\% of the variance of the evolution of the 90/10 achievement gap. The second model now includes the dummy of whether the country has an age selection of 15 or more and we can see that countries with late selection into tracking have about -.38 less standard deviations in inequality. Also note that the number of tracks dummy was now reduced, suggesting these two variables are explaining similar things. This is expected given that late tracking is usually associated with fewer tracks, such as in the Scandinavian countries. The third model now includes whether the country has any degree of tracking, meaning whether the country has above 0\% of the curriculum tracked. The coding of the variable comes from visual inspection of the distribution, where some countries, such as the United States and Canada, have untracked curriculums. For this last variable we see no relationship at all, but after including the vocational index, the model is completely reversed with countries with tracked curriculum having about 0.6 more standard deviations in inequality that countries with no tracking.

We also note that vocational enrollment is associated with less inequality. Not surprisingly, the age of selection coefficient, also increased due to the correlation between age of selection and vocational enrollment. Countries with high vocational enrollment are also those with early age selection, such as Germany and Austria. These two variables should be inspected further, perhaps with an interaction, given that both coefficients are sizable and very correlated. This final model explains over one third of the variance in the evolution of the 90/10 SES gap, with an R square of 40\%. The uncertainty of each of the covariates varies a lot with the percentage of curriculum tracked dummy ranging all the way from 0.9 to 0.3. Nonetheless, we see that the other coefficients are very strong and have reasonable intervals. It is very important to consider the uncertainty of our estimates, to gain confidence in our model fit. So far, this model seems to fit the data quite reasonably.
